---
title: 'Research Report: R'
author: "Valentin Jimenez"
date: "02/02/2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Third Part: Analysis with R

### Target question: **1.	How do annual members and casual riders use Cyclistic bikes differently?**

For this part of the project I decided to use R, because I had some ideas I could develop with python but I realized I could do the same with R, also to show how graphics look like with ggplot2 visualizations.

**1.** **Group by plane distance 12-Months**

For the distance traveled we can see from the graphics that patterns for both type of riders are pretty similar with no relevant differences.

```{r}
# In this script we analyze the number of riders, by type, taking into account the
# plane distance traveled.

library(ggplot2)
library(tidyverse)
library(readxl)


excelFile = read.csv("C:/Users/Valentín/Valasus Dropbox/Valentín Jiménez/PC/Documents/VJDS/Programming/coursera/googelDataAnalyticsCertificate/Course8/Track1_ CaseStudy1/bikes_2021/bigDatasetCSV.csv", header = TRUE,  na.strings=c(""," ","NA"))
#View(excelFile)

# We add a Months column to group_by after
excelFile <- excelFile %>% 
  mutate(Month = format(as.Date(excelFile$started_at),"%m"))
#View(excelFile)

# we calculate the total by type of member
total_by_memberType <- excelFile %>% 
  drop_na(start_lat,end_lat) %>%
  group_by(Month,member_casual) %>% 
  summarise(total_rider_type = n()) 

# transformation for graph
proof2 <-
  excelFile %>%
  select(rideable_type,started_at,start_lat,start_lng,end_lat,end_lng,
         member_casual,ride_duration_hours,Month) %>% 
  drop_na(start_lat,end_lat) %>% 
  mutate(plane_distance = sqrt((start_lat-end_lat)^2+(start_lng-end_lng)^2)) %>% 
  select(rideable_type, started_at, member_casual, ride_duration_hours,plane_distance,Month) %>% 
  mutate( Intervals = case_when
         ( plane_distance == 0 & ride_duration_hours == 0 ~ -1,
           plane_distance == 0 & ride_duration_hours > 0 ~ 0,
           plane_distance > 0 & plane_distance <= 0.01 ~ 0.01,
           plane_distance > 0.01 & plane_distance <= 0.02 ~ 0.02,
           plane_distance > 0.02 & plane_distance <= 0.03 ~ 0.03,
           plane_distance > 0.03 & plane_distance <= 0.04 ~ 0.04,
           plane_distance > 0.04 & plane_distance <= 0.05 ~ 0.05,
           plane_distance > 0.05 & plane_distance <= 0.06 ~ 0.06,
           plane_distance > 0.06 & plane_distance <= 0.07 ~ 0.07,
           plane_distance > 0.07 & plane_distance <= 0.08 ~ 0.08,
           plane_distance > 0.08 & plane_distance <= 0.09 ~ 0.09,
           plane_distance > 0.09 & plane_distance <= 0.1 ~ 0.1,
           TRUE ~ -1
         )
        ) %>% 
  group_by(Month,Intervals, member_casual) %>% 
  summarise(total_by_interval = n()) %>%
  group_by(Month, Intervals,member_casual) %>% 
  mutate(percentage = case_when(
      member_casual == "casual" ~ 
        as.numeric((total_by_interval*100)/total_by_memberType[as.numeric(Month)*2-1,3]),
      member_casual == "member" ~ 
        as.numeric((total_by_interval*100)/total_by_memberType[as.numeric(Month)*2,3])
    )
    ) 

#View(proof2)

# Here we separate the graph from the dataFrame which we name "proof2"
proof2 %>% 
  filter(Intervals != -1) %>% 
  group_by(Month) %>% 
  ggplot(aes(fill=member_casual, y = percentage , x = Intervals)) + 
  geom_bar(position="dodge", stat="identity")+
  theme(axis.text.x = element_text(angle = 45))+
  ggtitle("Percentage of Riders traveling a given distance")+
  labs(y = "Percentage by Type of Rider",caption = "12 months of 2021", 
       fill = "Type of Rider")+
  scale_x_continuous(breaks = seq(0,0.1,0.01),limits = c(0,0.1))+
  scale_fill_brewer(palette = "Paired" )+
  facet_wrap(~Month)



```

**2.** **Group by street annual**

In this part, first I considered the exact location of each station; however, I got so many values that I did not obtain relevant information, but if instead we group the stations by street name we get  more useful information: I used the initial name of a street to classify the stations. In the final results I show only names of stations with more or equal to 2% of users.

We see that annually, stations located at **Clark** street are by far the most used station for both type of riders. For casual riders there are specific streets used only by them, these are **Ashland, DuSable, Lake, Streeter**. Streets used by both riders but with casual riders being the most ones are **Michigan and Wells**. 

```{r}
# In this script we analyze the annual percentage of riders, by type, using stations
# located at a given street.

library(ggplot2)
library(tidyverse)
library(readxl)
library(stringr)

#excelFile = read.csv("C:/Users/Valentín/Valasus Dropbox/Valentín Jiménez/PC/Documents/VJDS/Programming/coursera/googelDataAnalyticsCertificate/Course8/Track1_ CaseStudy1/bikes_2021/bigDatasetCSV.csv", header = TRUE,  na.strings=c(""," ","NA"))
#View(excelFile)

# Here we use regex to form a column with the beginning of the name street so 
# we can calssify better
excelFile <- excelFile %>% 
  mutate(street = str_extract(string = excelFile$start_station_id_complete,
                              pattern = "^\\S+")) 

# we count the number of total riders by type using each street
excelFile <- excelFile %>% 
  drop_na(street) %>% 
  group_by(member_casual, street) %>% 
  summarise(n=n())

# we calculate the total by type of member and by street
totals_by_riderType <- excelFile %>% 
  group_by(member_casual) %>% 
  summarise(sum(n))

# we add the percentage column
excelFile <- excelFile %>%
  group_by(member_casual,street) %>% 
  mutate(percentage = case_when(
    member_casual == "casual" ~ as.numeric((n*100)/totals_by_riderType[1,2]),
    member_casual == "member" ~ as.numeric((n*100)/totals_by_riderType[2,2])
    ))
#View(excelFile)

# Now we can visualize properly
excelFile %>% 
  filter(percentage >=2) %>% 
  ggplot(aes(fill=member_casual, y = percentage , x = street)) + 
  geom_bar(position="dodge", stat="identity")+
  ggtitle("Annual percentage of Riders using stations located in a given Street")+
  labs(x = "Streets names", caption = "Year 2021", y = "Annual Percentage by Type of Rider",
       fill = "Type of rider")+
  scale_fill_brewer(palette = "RdPu")+
  theme(axis.text.x = element_text(angle = 45))
```

**3.** **Group by street 12-Months**

This part follows the ideas of the point **2.** of this part of the project, but now we look for street names at each Month of 2021 and with a monthly percentage of more or equal to 2.5%.

We confirm that for each month **Clark** is still the most used one for both type of riders. We can see that the streets **Broadway, Lake, Wells**, are used by casual members through out each month of the year.

```{r}
# In this script we analyze the number of riders, using the stations
# located at a certain street.

library(ggplot2)
library(tidyverse)
library(readxl)
library(stringr)

excelFile = read.csv("C:/Users/Valentín/Valasus Dropbox/Valentín Jiménez/PC/Documents/VJDS/Programming/coursera/googelDataAnalyticsCertificate/Course8/Track1_ CaseStudy1/bikes_2021/bigDatasetCSV.csv", header = TRUE,  na.strings=c(""," ","NA"))
#View(excelFile)

# Here we use regex to form a column with the beginning of the name street so 
# we can calssify better
excelFile <- excelFile %>% 
  mutate(street = str_extract(string = excelFile$start_station_id_complete,
                              pattern = "^\\S+")) %>% 
  mutate(Month = format(as.Date(excelFile$started_at),"%m")) %>% 
  drop_na(street) %>% 
  group_by(Month,member_casual,street) %>% 
  summarise(n=n()) 
#View(excelFile)

# we calculate the total by type of rider and by month
totals_by_Month <- excelFile %>% 
  group_by(Month,member_casual) %>% 
  summarise(sum_Month = sum(n))
#View(totals_by_Month)

# we add the percentage column
excelFile <- excelFile %>%
  group_by(Month, member_casual,street) %>% 
  mutate(percentage = case_when(
    member_casual == "casual" ~ as.numeric((n*100)/totals_by_Month[as.numeric(Month)*2-1,3]),
    member_casual == "member" ~ as.numeric((n*100)/totals_by_Month[as.numeric(Month)*2,3])
    )) 
#View(excelFile)

# Now we can visualize properly
  excelFile %>% 
  #select(Month, member_casual,street,percentage) %>% 
  filter(percentage >2.5) %>% 
  ggplot(aes(fill=member_casual, y = percentage , x = street)) + 
  geom_bar(position="dodge", stat="identity")+
  theme(axis.text.x = element_text(angle = 45))+
  ggtitle("Percentage of Riders using the stations belonging to the given street")+
  labs(subtitle = "Showing only percentage > 2.5", y = "Percentage by Month", x = "Streets",
       fill = "Type of Rider",caption = "12 Months of 2021")+
  scale_fill_brewer(palette = "Paired")+
  facet_wrap(~Month, ncol = 3)

```






